---
title: "Ich data"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

# Evaluation of YBFMP ichthyoplankton data

## Background:

-   The collection of fish larvae and eggs measures seasonal variation in densities and species assemblages within the Yolo Bypass. Sampling began in 1999 and occurs biweekly from January through July and used to occur year-round. The net used consists of 500-micron mesh and is 0.65m in diameter and 2m long.

    -   1999-2010: sampling at rotary screw trap site (STTD) and Sacramento River at Sherwood Harbor (SHR) at least once/month from March-June; sometimes started in January or collected weekly during inundation and drainage events.

    -   2011-2015: Sampling at least every other week year-round at STTD and SHR (weekly during inundation).

    -   2016: Sampling at SHR and STTD reduced to Jan-July due to low catch during July-Dec.

    -   2017: Sampling suspended at SHR due to low/lack of catch.

    -   2018: Samples collected at STTD year-round.

    -   2019: Sampling at STTD returned to Jan-July.

    -   Mar - Oct 2020, Dec 2020 - Apr 2021 : sampling suspended due to COVID-19.

-   The collection of ichthyoplankton samples is an important element in determining the annual presence, timing, and recruitment success of fishes utilizing the Yolo Bypass.

    -   Key findings to date include:

        1.  26 species of fish larvae were observed in the Yolo Bypass during the 20-years of monitoring (Mollie Ogaz and J. Frantzich, DWR, unpublished data), including Delta Smelt, *Hypomesus transpacificus* (Sommer et al 2004b);

        2.  The native Prickly Sculpin and non-native Threadfin Shad dominated samples, contributing to over 60% of the total larval catch (Mollie Ogaz and J. Frantzich, DWR, unpublished data);

        3.  Native species made up a higher percent of total catch in the Yolo Bypass (22.43%) in comparison to in the Sacramento River (10.2%), and appeared earlier in the year than many non-natives (Mollie Ogaz and J. Frantzich, DWR, unpublished data);

        4.  Similar to other seasonal floodplains in the San Francisco Estuary, alien fishes made up a large portion of the assemblage of early life stages in the Yolo Bypass (Sommer et al 2004b);

        5.  Water temperature and stage are the best explanatory environmental variables for larval fish abundance in the Yolo Bypass (p=0.001). Flow was not statistically significant (Mollie Ogaz and J. Frantzich, DWR, unpublished data);

        6.  Species richness and diversity are higher in the Yolo bypass in comparison to in the Sacramento River (Sommer et al. 2004b).

## Current Data Collection Practices:

### YBFMP

-   Currently, samples are collected every 2 weeks from January - June at STTD.

    -   Sample frequency increases to every week during bypass inundation.

### Other IEP Surveys

-   20-mm: Fish samples are collected using an egg and larval, rigid-mouthed net constructed of 1,600 µm mesh.

-   DJFMP Liberty Island work (concluded): 500 µm larval net is used to collect larval fish.

-   Smelt Larva Survey: At each site, a single 10-minute oblique tow is conducted using a 500-micron mesh egg and larval net.

## YBFMP Ichthyoplankton Data Overview

### Overall catch

WARNING - data displayed as count only at this time, not standardized for effort!

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate) 

ichcatch <- read_csv("FullData.csv")

ichcatch$Date <- as.Date(ichcatch$Date, format = "%Y-%m-%d")
ichcatch$Year <- as.integer(format(ichcatch$Date, format = "%Y"))

#sum by date

#shift between 4/15 and 4/22/2019 in terms of how count is recorded - adjust so we can look at it all together
ich.day.old <- ichcatch %>% 
  filter(Date < "2019-04-16") %>%
  group_by(Year, MonthAbb, Date, Station, ScientificName) %>%
  summarize("Tot.Count"= sum(Count, na.rm = T))

ich.day.new <- ichcatch %>% 
  filter(Date > "2019-04-16") %>%
  group_by(Year, MonthAbb, Date, Station, ScientificName) %>%
  summarize("Tot.Count"= mean(TotalCountSpecies, na.rm = T))

ich.day.combo <- rbind(ich.day.new, ich.day.old)

#all data
ich.day.combo %>%
   ggplot(aes(x = Date, y = Tot.Count, color=ScientificName)) +
    geom_point(alpha=0.6) + 
    facet_wrap(~Station, ncol = 1, scales = "free_y") +
    theme_bw() +
    labs(x = "Date", y= "Count") +
    theme(legend.position="none")
```

### Annual catch

```{r, message=FALSE, warning=FALSE}
#by year
ich.day.combo.yr <- ich.day.combo %>% 
  group_by(Year, Station, ScientificName) %>%
  summarize("Yr.Count"= sum(Tot.Count, na.rm = T))

ich.day.combo.yr %>%
   ggplot(aes(x = Year, y = Yr.Count, fill=ScientificName)) +
    geom_bar(stat='identity', position='stack') + 
    facet_wrap(~Station, ncol = 1, scales = "free_y") +
    theme_bw() +
    labs(x = "Year", y= "Total Count") +
     theme(legend.position="none")

pal <- c("#004949","#009292","#ff6db6","#ffb6db","#490092","#006ddb","#b66dff","#6db6ff","#920000","#924900","#b6dbff","#db6d00","#24ff24","#ffff6d", "#465177","#E4C22B","#965127","#29483A","#759C44","#9FB6DA","#DF3383")
 
#STTD 2011-2022
ich.day.combo.yr %>%
  filter(Station=="STTD") %>%
   filter(Year>"2011") %>%
    filter(Yr.Count>10) %>%
   ggplot(aes(x = Year, y = Yr.Count, fill=ScientificName)) +
    geom_bar(stat='identity', position='stack', color="black") + 
    theme_bw() +
    scale_fill_manual(values=pal) +
    labs(x = "Year", y= "Total Count")
#decrease from 2015 to 2016 partly due to decreased sampling
```

### Monthly catch

```{r, message=FALSE, warning=FALSE}
#STTD by month
ich.day.combo.mo <- ich.day.combo %>% 
  group_by(Year, MonthAbb, Station, ScientificName) %>%
  summarize("Mo.Count"= sum(Tot.Count))

ich.day.combo.mo$MonthAbb <- factor(ich.day.combo.mo$MonthAbb, levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))

ich.day.combo.mo %>%
  filter(Station=="STTD") %>%
  filter(Year %in% c("2011", "2012", "2013", "2014", "2015")) %>%
   ggplot(aes(x = MonthAbb, y = Mo.Count, fill=ScientificName)) +
    geom_bar(stat='identity', position='stack', color="black") + 
    facet_wrap(~Year, ncol = 1, scales = "free_y") +
    theme_bw() +
      scale_fill_manual(values=pal) +
    labs(x = "Month", y= "Total Count") 

#STTD by month 2011+
ich.day.combo.mo %>%
  filter(Station=="STTD") %>%
  filter(Mo.Count>10) %>%
  filter(Year>2010) %>%
  filter(MonthAbb %in% c("Jan", "Feb", "Mar", "Apr", "May", "Jun")) %>%
   ggplot(aes(x = MonthAbb, y = Mo.Count, fill=ScientificName)) +
    geom_bar(stat='identity', position='stack') + 
    theme_bw() +
        scale_fill_manual(values=pal) +
    labs(x = "Month", y= "Total Count") 

#STTD by month 2011+, cyprinids
ich.day.combo.mo %>%
  filter(Station=="STTD") %>%
  filter(Year>2010) %>%
  filter(ScientificName %in% c("Cyprinidae", "Pogonichthys macrolepidotus", "Hypomesus transpacificus
", "Acipenser transmontanus", "Orthodon microlepidotus")) %>%
  filter(MonthAbb %in% c("Jan", "Feb", "Mar", "Apr", "May", "Jun")) %>%
   ggplot(aes(x = MonthAbb, y = Mo.Count, fill=ScientificName)) +
    geom_bar(stat='identity', position='stack') + 
    theme_bw() +
        scale_fill_manual(values=pal) +
    labs(x = "Month", y= "Total Count") 
```

Thoughts:

-   During years where sampling occurred at least twice a month year-round at STTD, the highest counts occurred Feb-Aug

-   Prickly sculpin show up Feb-Apr, Threadfin shad Jun-Aug, American shad May-Jul

-   Over the last 12 years, when we do catch Cyprinids its Mar-May with the most identifiable as Splittail in April

-   Blackfish have only been identified in May

-   Data gives us an idea of which fish use the bypass or upper Cache Slough complex for spawning/early recruitment, however:

    -   no comparison with Sac

    -   only decent numbers of a few species (shads, sculpin, gobies)

## Recommendation:

1.  Continue current sampling regime.

    -   January and June contain few native species but are valuable as bookends as conditions change due to climate and restoration (Big Notch especially).

    -   Later summer months do see nonnative larval fishes, but adding sampling for species of little management concern would not be the most effective use of program resources.
